---
title: "Validation with omics data"
date: "`r Sys.Date()`"
output: word_document
---

It is assumed that `analyze_slimm_t2d_omics.Rmd` was already run, so `ezlimma` and `Hitman` have been installed. Then, installation time is under one minute. This code was all run on a desktop.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=FALSE)
options(digits=3, stringsAsFactors=FALSE)
```

```{bash}
set DISPLAY=
# ssh ...
srun -c 27 -p priority --pty -t 1-0 /bin/bash 
module load R
R
```

We validate our methods similarly to Barfied et al. (2017) except we simulate omics datasets with hundreds of analytes or genes whose significance is assessed with the false discovery rate (FDR) rather than with p-values.

```{r params}
setwd("/n/jobspace/bbcore/slimm-t2d-omics/validation")

# library(remotes)
# remotes::install_github("jdreyf/Hitman")
library(Hitman)

nsim <- 10**3
nsamp.v <- c(15, 50)
fdr <- 0.15
ngene <- 500
prop.v <- c(1, 5, 25)/ngene
# scenario
sc.v <- paste0("sc", 1:3)
b1 <- t2 <- 2
# with prop.1c<=20% I see almost no messages that E not assoc w/ Y
prop.em <- 0.4
sd.mn <- 0

t1.nm <- "5x"

cov.type <- "gtex"
if (cov.type == "structured"){
  cov_mat <- matrix(0.3, nrow=ngene, ncol=ngene)
  diag(cov_mat) <- 1
  Sigma.sc1 <- diag(ngene)
  cov.type.out <- "Str"
} else {
  cov_mat <- as.matrix(read.csv("../data/gtex_muscle_cov_subset_scaled.csv", row.names = 1))
  Sigma.sc1 <- matrix(0, nrow=nrow(cov_mat), ncol=ncol(cov_mat))
  diag(Sigma.sc1) <- diag(cov_mat)
  cov.type.out <- "GTEx"
}
```


## Simulate scenerios


Scenario 1: Genes are independent and mediators are consistent

Scenario 2: Genes are dependent and mediators are consistent

Scenario 3: Genes are dependent and there are both consistent and inconsistent mediators

```{r sim, eval=FALSE}
library(foreach)
library(doParallel)

registerDoParallel(cores=27)

res <- foreach(nsamp = nsamp.v) %:%
  foreach(prop = prop.v) %:%
    foreach(sc = sc.v, .packages='Hitman') %dopar% {
      if (sc=="sc1"){
        t1 <- 5*prop*ngene*b1
        sc1 <- Hitman::sim_omics(nsamp=nsamp, ngene=ngene, FDR=fdr, prop.consistent=prop, b1 = b1, t2=t2, t1=t1, nsim=nsim,
                                  Sigma = Sigma.sc1, prop.em = prop.em)
        write.csv(sc1, paste0("sim_omics_sc1_nsim", nsim,  "_prop", prop, "_t1.", t1.nm, "_ngene", ngene, "_N", nsamp, 
                              "_b1t2", b1, "_propem", prop.em, "_cov", cov.type.out, ".csv"))
        sc1
      } else if (sc=="sc2"){
        t1 <- 5*prop*ngene*b1
        sc2 <- Hitman::sim_omics(nsamp=nsamp, ngene=ngene, FDR=fdr, prop.consistent=prop, b1 = b1, t2=t2, t1=t1, nsim=nsim, 
                                  Sigma = cov_mat, prop.em = prop.em)
        write.csv(sc2, paste0("sim_omics_sc2_nsim", nsim, "_prop", prop, "_t1.", t1.nm, "_ngene", ngene, "_N", nsamp,  
                              "_b1t2", b1, "_propem", prop.em, "_cov", cov.type.out, ".csv"))
        sc2
      } else if (sc=="sc3"){
        t1 <- 10*prop*ngene*b1
        # prop2 <- ceiling(prop*ngene/2)/ngene
        sc3 <- Hitman::sim_omics(nsamp=nsamp, ngene=ngene, FDR=fdr, prop.consistent=prop, prop.inconsistent = prop,
                                       b1 = b1, t2=t2, t1=t1, nsim=nsim, Sigma = cov_mat, prop.em = prop.em)
        write.csv(sc3, paste0("sim_omics_sc3_nsim", nsim, "_prop", prop, "_t1.", t1.nm, "_ngene", ngene, "_N", nsamp,  
                              "_b1t2", b1, "_propem", prop.em, "_cov", cov.type.out, ".csv"))
        sc3
      }
    }

# stopImplicitCluster()
```


## Compile tables


Compile output from our simulations into one table per scenario.

```{r collect}
sc1 <- sc2 <- data.frame(Method=rep(c("Hitman", "Lotman", "Joint signif"), each=length(prop.v)*length(nsamp.v)), 
                         N=rep(nsamp.v, each=length(prop.v)), prop=rep(prop.v, times=length(nsamp.v)), 
                         N_mediators=ngene*rep(prop.v, times=length(nsamp.v)), N_signif=NA, Power=NA, FDR=NA)

sc3 <- data.frame(Method=rep(c("Hitman", "Lotman", "Joint signif consistent", "Joint signif both"), 
                             each=length(prop.v)*length(nsamp.v)), 
                  N=rep(nsamp.v, each=length(prop.v)), prop=rep(prop.v, times=length(nsamp.v)), 
                  N_cons_med=ngene*rep(prop.v, times=length(nsamp.v)), 
                  N_incons_med=ngene*rep(prop.v, times=length(nsamp.v)),
                  N_signif=NA, Power=NA, FDR=NA)

for (nsamp in nsamp.v){
  for (prop in prop.v){
    sc1.tmp <- read.csv(paste0("sim_omics_sc1_nsim", nsim,  "_prop", prop, "_t1.", t1.nm, "_ngene", ngene, 
                               "_N", nsamp, "_b1t2", b1, "_propem", prop.em, "_cov", cov.type.out, ".csv"))
    sc1.rnames <- rownames(sc1)[sc1$N == nsamp & sc1$prop == prop]
    sc1[sc1.rnames, 5:7] <- sc1.tmp[, c("n.hits", "power", "fdr")]
  
    sc2.tmp <- read.csv(paste0("sim_omics_sc2_nsim", nsim,  "_prop", prop, "_t1.", t1.nm, "_ngene", ngene, 
                               "_N", nsamp, "_b1t2", b1, "_propem", prop.em, "_cov", cov.type.out, ".csv"))
    sc2.rnames <- rownames(sc2)[sc2$N == nsamp & sc2$prop == prop]
    sc2[sc2.rnames, 5:7] <- sc2.tmp[, c("n.hits", "power", "fdr")]
    
    sc3.tmp <- read.csv(paste0("sim_omics_sc3_nsim", nsim,  "_prop", prop, "_t1.", t1.nm, "_ngene", ngene, 
                               "_N", nsamp, "_b1t2", b1, "_propem", prop.em, "_cov", cov.type.out, ".csv"))
    sc3.rnames <- rownames(sc3)[sc3$N == nsamp & sc3$prop == prop]
    sc3[sc3.rnames, 6:8] <- sc3.tmp[, c("n.hits", "power", "fdr")]
  }
}
```

Write out a table per scenario.

```{r write}
sc1.wr <- sc1[order(sc1$N, sc1$N_mediators), setdiff(colnames(sc1), "prop")]
write.csv(sc1.wr, paste0("combined_sc1_omics_nsim", nsim, "_b1t2", b1, "_propem", prop.em, "_cov", cov.type.out, ".csv"), row.names = FALSE)

sc2.wr <- sc2[order(sc2$N, sc2$N_mediators), setdiff(colnames(sc2), "prop")]
write.csv(sc2.wr, paste0("combined_sc2_omics_nsim", nsim, "_b1t2", b1, "_propem", prop.em, "_cov", cov.type.out, ".csv"), row.names = FALSE)

sc3.wr <- sc3[order(sc3$N, sc3$N_cons_med), setdiff(colnames(sc3), "prop")]
write.csv(sc3.wr, paste0("combined_sc3_omics_nsim", nsim, "_b1t2", b1, "_propem", prop.em, "_cov", cov.type.out, ".csv"), row.names = FALSE)
```

Test differences in power when only one mediator, since then power * number simulations is binomial.

```{r ttest}
sc1.wr <- read.csv(paste0("validation/combined_sc1_omics_nsim", nsim, "_b1t2", b1, "_propem", prop.em, "_cov", 
                          cov.type.out, ".csv"))

hit.prop <- nsim*signif(sc1.wr %>% 
                               filter(Method=="Hitman", N_mediators==1, N==15) %>% 
                               select("Power"), 5)[1, 1]
js.prop <- nsim*signif(sc1.wr %>% 
                         filter(Method=="Joint signif", N_mediators==1, N==15) %>% 
                         select("Power"), 5)[1, 1]

tt.n15 <- t.test(x=rep(1:0, c(hit.prop, nsim-hit.prop)),
                y=rep(1:0, nsim*c(js.prop, nsim-js.prop)),
                var.equal = TRUE)
print(tt.n15)
```


# References
Barfield R, Shen J, Just AC, Vokonas PS, Schwartz J, Baccarelli AA, VanderWeele TJ, Lin X. Testing for the indirect effect under the null for genome-wide mediation analyses. Genet Epidemiol. 2017 Dec;41(8):824-833.
