---
title: "Validations by simulation"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=FALSE)
options(digits=3, stringsAsFactors=FALSE)
library(ezlimma)
```

We validate the ability of each of our methods (`hitman`, `lotman`, and `pants`) to control their false positive rate (*size*) and have high rate of detecting true positives (*power*) via simulations, where we know the truth.

# Mediation

We simulate our mediation methods using the benchmark simulation from Barfield et al. (2017).

## Lotman

This calculation takes ~10 min on my desktop PC. Our simulation results are at [sim_lotman_nsamp50_nsim10000.csv](./results/sim_lotman_nsamp50_nsim10000.csv).

```{r, eval=FALSE}
prop.sig.mat <- ezlimma:::sim_barfield(med.fcn = lotman, b1t2.v=c(0, 0.14, 0.39), nsim = 10**4)
write.csv(prop.sig.mat, "./results/sim_lotman_nsamp50_nsim10000.csv", na="")
```

## Hitman

This calculation takes ~30 min on my desktop PC. Our simulation results are at [sim_hitman_nsamp50_nsim10000.csv](./results/sim_hitman_nsamp50_nsim10000.csv).

```{r, eval=FALSE}
prop.sig.mat <- ezlimma:::sim_barfield(med.fcn = hitman, b1t2.v=c(0, 0.14, 0.39), nsim = 10**4, ngene=99)
write.csv(prop.sig.mat, "./results/sim_hitman_nsamp50_nsim10000.csv", na="")
```

# Pants



# References
Barfield R, Shen J, Just AC, Vokonas PS, Schwartz J, Baccarelli AA, VanderWeele TJ, Lin X. Testing for the indirect 
effect under the null for genome-wide mediation analyses. Genet Epidemiol. 2017 Dec;41(8):824-833.